---
title: "HMP16SData"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
author:
    - Lucas Schiffer
    - Rimsha Azhar
    - Marcel Ramos
    - Ludwig Geistlinger
    - Levi Waldron
package: HMP16SData
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{HMP16SData}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
---

# Prerequisites

```{r, message=FALSE}
BiocInstaller::biocLite("Bioconductor/SummarizedExperiment")
BiocInstaller::biocLite("schifferl/kableOne")

library(HMP16SData)
library(magrittr)
# library(phyloseq)
library(kableOne)
```

# Introduction

`r BiocStyle::Biocpkg("HMP16SData")` is a Bioconductor ExperimentHub package of
the Human Microbiome Project (HMP) 16S ribosomal RNA (rRNA) sequencing data for
variable regions 1-3 and 3-5. The raw data files are provided in the package as
downloaded from the HMP Data Analysis and Coordination Center (DACC) website. 
Processed data is provided as SummarizedExperiment objects via ExperimentHub.

`r BiocStyle::Biocpkg("HMP16SData")` can be installed using 
`r BiocStyle::Biocpkg("BiocInstaller")` as follows:

```{r, eval=FALSE}
BiocInstaller::biocLite("HMP16SData")
```

Once installed, `r BiocStyle::Biocpkg("HMP16SData")` provides two functions to 
access data – one for variable region 1-3 and another for variable region 3-5. 
When called as follows the functions will download data from an 
`r BiocStyle::Biocpkg("ExperimentHub")` Amazon S3 (Simple Storage Service) 
bucket over https.

```{r, message=FALSE}
V13()
V35()
```

The two datasets are represented as `SummarizedExperiment` objects, a standard 
Bioconductor class that is amenable to subsetting and analysis. To maintain 
breveity, details of the `SummarizedExperiment` class are not outlined here but 
the `r BiocStyle::Biocpkg("SummarizedExperiment")` package provides an excellent
vignette.

# Features

## Straightforward Subsetting

```{r, message=FALSE}
V13_stool <-
    V13() %>%
    subset(select = hmp_body_subsite == "Stool")

V13_stool
```

## Protected Metadata Attachment

An example is shown here and there is an additional vignette on obtaining 
protected metadata.

```{r, eval=FALSE}
V13_stool %>%
    attach_dbGaP("~/ncbi/dbGaP-12146/", pretty_colnames = TRUE) %>%
    as_phyloseq() %>%
    plot_richness(x = "diet", color = "diet", measures = "Shannon")
```

## Taxonomic Identifier Translation

```{r, eval=FALSE}
V13_stool %>%
    translate_identifiers()
```

## External Data Comparison

Topic to be covered in a searate vignette.

## Frequency Table Generation

Be aware that the `kable_one` function from the
`r BiocStyle::Githubpkg("schifferl/kableOne")` package is _very_ beta.

```{r, message=FALSE}
V13 <-
    V13() %>%
    table_one()

V35 <-
    V35() %>%
    table_one()

kable_one(V13, V35)
```

# Limitations

Data from variable regions 1-3 and 3-5 are not comparable to everything.
