---
title: "HMP16SData"
date: '`r format(Sys.Date(), "%B %e, %Y")`'
author:
    - Lucas Schiffer
    - Rimsha Azhar
    - Marcel Ramos
    - Ludwig Geistlinger
    - Levi Waldron
output:
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{HMP16SData}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
---

# Introduction

`r BiocStyle::Biocpkg("HMP16SData")` is a Bioconductor ExperimentHub package of 
the Human Microbiome Project (HMP) 16S rRNA sequencing data for variable regions
1-3 and 3-5. The raw data files are provided in the package as downloaded from 
the HMP Data Analysis and Coordination Center website. Processed data is 
provided as SummarizedExperiment objects via ExperimentHub.

`r BiocStyle::Biocpkg("HMP16SData")` can be installed using 
`r BiocStyle::Biocpkg("BiocInstaller")` as follows:

```{r, eval=FALSE}
BiocInstaller::biocLite("HMP16SData")
```

Once installed, `r BiocStyle::Biocpkg("HMP16SData")` provides two functions to 
access data – one for variable region 1-3 and another for variable region 3-5. 
When called as follows the functions will download data from an 
`r BiocStyle::Biocpkg("ExperimentHub")` AWS (Amazon Web Services) S3 (Simple 
Storage Service) bucket over https. Of couse, it is necessary to load the 
package first.

```{r, message=FALSE, results='hold'}
library(HMP16SData)
V13()
V35()
```

The two datasets are represented as `SummarizedExperiment` objects, a standard 
Bioconductor class that is amenable to subsetting and analysis. To maintain 
breveity, details of the `SummarizedExperiment` class are not outlined here but 
the `r BiocStyle::Biocpkg("SummarizedExperiment")` package provides an excellent
vignette.

# Features

## Attach Protected Metadata from dbGAP

## Subet Data by Bodysite

## Translate Taxonomic Identifiers

# Analysis

## Import External Data

## Convert to Relative Abundance

## Coercing to a `r BiocStyle::Biocpkg("phyloseq")` object


